Unsupervised Learning for Identifying Events in Active Target Experiments

### 2020-08-05 doc/Manuscript/article.tex
Updates from Overleaf

The concept of active targets aim at mitigating this compromise, by turning the target itself into a detector \cite{BECEIRONOVO2015}. Most active target detectors such as the AT-TPC are composed of a time projection chamber (TPC) where the detector gas is at the same time the target material. Recoil particles that originate from a nuclear reaction between a beam nucleus and a gas nucleus can be tracked from the vertex of the reaction to their final position inside the active volume of the target. Their properties can therefore be measured without any loss of resolution regardless of the amount of material traversed by the beam. At the same time, the detection efficiency is dramatically increased by the very large solid angle covered in this geometry. A direct consequence of this concept is the inclusiveness of the experimental data recorded by this type of detector: any nuclear reaction happening within the target is recorded. Although this sounds like an advantage from the scientific point of view, it poses great challenges during the analysis phase, that are reminiscent of bubble chamber times and on par with event classification challenges in particle physics today, see for example the recent review of Mehta {\em et al.} \cite{mehta2019}. More often than not, the reaction channel of interest has one of the lowest cross sections. When analyzing the data one is therefore faced with the task of sorting out the corresponding events from the {\em background} of other reaction channels.

% motivating ML work:
Because TPCs produce three-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (comparable to the analyses made in the bubble chamber era), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently \cite{mehta2019}. In addition, machine learning (ML) algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled data set is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know {\em a priori} the types and behaviors of the reactions present in the detector in order to generate a labeled and simulated data set. In the latter case, there must also exist a validation data set of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

@@

The AT-TPC data we studied for this work was recorded as charge time-series for each of the  the $\sim10^4$ detector pads.
In this representation, an event is a record of $512$ time-buckets for each of the $10^4$ detector pads. In our analysis, we represent each event as a down-sampled two-dimensional  projection.
We chose to represent the data in two dimensions to facilitate the use of sophisticated image-recognition machine learning models, and this data representation was shown to successfully classify the events of this experiment in a supervised setting \cite{Kuchera2019}.
First, the time-series data was represented as a three-dimensional  point cloud, where each point contains the maximum charge in the time-series trace. We log-transform then min-max scale the charge, mapping the charge to the interval $[0, 1]$. The data is projected into two-dimensional  space by summing over the time-axis, then down-sampled into a $128\times128$ pixel image.
%Additionally, we log-transform and normalize the data to the interval $[0, 1]$ for each point. The final representation for each event is a $128 \times 128$ matrix with each element ranging from $0$ to $1$.

One of the significant considerations for the analysis of AT-TPC data is to inject machine learning methods for track identification at the best point in the analysis pipeline.
Using raw data is advantageous as it provides an unbiased view of the event, but the data volume and noise levels might be prohibitive for analysis.
%On the opposite end of the pipeline, 
Therefore, we incrementally add bias to the analysis by applying the algorithm further down the analysis pipeline, with the benefit being that more preprocessing improves the signal-to-noise ratio, possibly improving model performance.
To explore this trade-off between model performance and preprocessing bias, we performed our analysis on simulated, cleaned and raw events.

\begin{table}
\centering
\caption{Event class percentages for each of the data sets used in this analysis. The decrease in the "other" class of events from full to filtered data owes to the thresholding of events. If an event contains fewer than $20$ data points it is discarded }\label{tab:class_distr}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
$\%$ Proton & $50$ & $25.3$ & $27.6$ \\ 
$\%$ Carbon & $50$ & $12.0$ & $11.9$ \\
$\%$ Other & $0$ & $62.8$ & $60.1$ \\
\end{tabular}
\end{table}

@@

A set of $N=40000$ simulated AT-TPC events per class was obtained  from a ${}^{46}$Ar$(p, p)$ experiment with the \lstinline{pytpc} package developed by \citet{Bradt2017a}. %Using the same parameters {\color{blue} what parameters are you referring to? -MPK} as for the 
%Two simulated ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.

>>>>

The concept of active targets aim at mitigating this compromise, by turning the target itself into a detector \cite{BECEIRONOVO2015}. Most active target detectors such as the AT-TPC are composed of a time projection chamber (TPC) where the detector gas is at the same time the target material. Recoil particles that originate from a nuclear reaction between a beam nucleus and a gas nucleus can be tracked from the vertex of the reaction to their final position inside the active volume of the target. Their properties can therefore be measured without any loss of resolution regardless of the amount of material traversed by the beam. At the same time, the detection efficiency is dramatically increased by the very large solid angle covered in this geometry. A direct consequence of this concept is the inclusiveness of the experimental data recorded by this type of detector: any nuclear reaction happening within the target is recorded. Although this sounds like an advantage from the scientific point of view, it poses great challenges during the analysis phase, that are reminiscent 
of bubble-chamber times \cite{Hough:1959} and on par with event classification challenges in particle physics today, see for example the recent review of Mehta {\em et al.} \cite{mehta2019}. More often than not, the reaction channel of interest has one of the lowest cross sections. When analyzing the data one is therefore faced with the task of sorting out the corresponding events from the {\em background} of other reaction channels.

% motivating ML work:
Because TPCs produce three-dimensional images of charged particle tracks, the event identification task is often akin to a visual inspection (comparable to the analyses made in the bubble-chamber era \cite{Hough:1959}), which is not practical nowadays because of the large quantities of data. Machine learning techniques then appear as a promising prospect, in particular in the image recognition domain where much progress has been made recently \cite{mehta2019}. In addition, machine learning (ML) algorithms offer new possibilities such as the potential discovery of unforeseen phenomena that would have been missed by more traditional analysis methods. Prior work has demonstrated the ability to apply supervised classification machine learning methods to AT-TPC data when a labeled training set is available, whether through hand-labeled data or labeled simulated data (a {\em transfer learning} application) \cite{Kuchera2019}. In some experiments, a labeled data set is unavailable. This could be  due to the inability to hand-label events, or the case where one does not know {\em a priori} the types and behaviors of the reactions present in the detector in order to generate a labeled and simulated data set. In the latter case, there must also exist a validation data set of real data, still requiring the ability to label a subset of the real data. The unsupervised separation of event types, or {\em clustering}, based on a set of ML algorithms is hereby examined, using experimental data recorded by the AT-TPC during its commissioning experiment from a radioactive $^{46}$Ar beam reacting on an isobutane target composed of proton and carbon nuclei.

@@

The AT-TPC data we studied for this work was recorded as charge time-series for each of the  the $\sim10^4$ detector pads.
In this representation, an event is a record of $512$ time-buckets for each of the $10^4$ detector pads. In our analysis, we represent each event as a down-sampled two-dimensional  projection.
We chose to represent the data in two dimensions to facilitate the use of advanced image-recognition machine learning models, and this data representation was shown to successfully classify the events of this experiment in a supervised setting \cite{Kuchera2019}.
First, the time-series data was represented as a three-dimensional  point cloud, where each point contains the maximum charge in the time-series trace. We log-transform and map the data to the interval $[0, 1]$. The data is projected into a two-dimensional  space by summing over the time-axis. Thereafter the data is down-sampled into a $128\times128$ pixel image.
%Additionally, we log-transform and normalize the data to the interval $[0, 1]$ for each point. The final representation for each event is a $128 \times 128$ matrix with each element ranging from $0$ to $1$.

One of the significant considerations for the analysis of AT-TPC data is to inject machine learning methods for track identification at the best point in the analysis pipeline.
Using raw data is advantageous as it provides an unbiased view of the event, but the data volume and noise levels might be prohibitive for the analysis.
%On the opposite end of the pipeline, 
Therefore, we incrementally add bias to the analysis by applying the algorithm further down the analysis pipeline, with the benefit being that more preprocessing improves the signal-to-noise ratio, possibly improving model performance.
To explore this trade-off between model performance and preprocessing bias, we performed our analysis on simulated, cleaned and raw events as discussed below.

\begin{table}
\centering
\caption{Event class percentages for each of the data sets used in this analysis. The decrease in the "other" class of events from full to filtered data owes to the thresholding of events. If an event contains fewer than $20$ data points it is discarded }\label{tab:class_distr}
\begin{tabular}{lccc}
\toprule
{} & Simulated & Full & Filtered \\
\midrule
Percentage Proton & $50$ & $25.3$ & $27.6$ \\ 
Percentage Carbon & $50$ & $12.0$ & $11.9$ \\
Percentage Other & $0$ & $62.8$ & $60.1$ \\
\end{tabular}
\end{table}

@@

A set of $N=40000$ simulated AT-TPC events per class was obtained  from a ${}^{46}$Ar$(p, p)$ experiment with the \lstinline{pytpc} package developed by \citet{Bradt2017a}, see table \ref{tab:class_distr}. %Using the same parameters {\color{blue} what parameters are you referring to? -MPK} as for the 
%Two simulated ${}^{46}$Ar$(p, p)$ experiment, a small set of $N=4000$ events per class was generated, as well as a larger set of $N=40000$ events per class.
